---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
##### Author: Andria Hall

##### Date: "March 15, 2015"

## Introduction

This report is a **Reproducible Research** and is a **single R markdown** document that is processed by `knitr` and transformed into a `HTML` file. The dataset is stored in a comma-separated-value (CSV) file and there are a total of **17,568** observations in this dataset. 

The variables included in this dataset are:

* **steps:** Number of steps taking in a 5-minute interval (missing values are
coded as NA)

* **date:** The date on which the measurement was taken in YYYY-MM-DD
format.

* **interval:** Identifier for the 5-minute interval in which measurement was
taken.

The dataset is generated from a personal activity monitoring device that collects data at a 5-minute interval, throughout the day, as stated above, and by an anonymous individual during the months of **October** and **November 2012**.


## Prepare the working environment

Since the report requires writing code chunks in the **R markdown** document, so that my peer evaluators will be able to review the code for their analysis; I have set `echo = TRUE` and  `results = hold` as global options.

```{r load_knitr_library}
# load the knitr library
library(knitr)
opts_chunk$set(echo = TRUE, results = 'hold')
```

# Load required packages

```{r load_packages}
# load required packages
suppressMessages(require("data.table"))
suppressMessages(require("dplyr"))
suppressMessages(require("tidyr"))
suppressMessages(require("ggplot2"))
```

## Loading and preprocessing the data

Here the activity dataset is loaded and processed for analysis.

```{r load_activity}
# load the activity data
setwd("/Users/Andria/RepData_PeerAssessment1")
f <- file.path(getwd(), "activity.zip")
activity <- tbl_df(read.csv(unz(f, "activity.csv"),header= TRUE, sep= ","))

# explore activity data
head(select(activity, steps:interval))
```

## What is mean total number of steps taken per day?

Here the missing `NA` values are ignored and the **total** number of steps are summarised.

```{r compute_steps_taken}
# calculate total steps taken
steps_taken <- aggregate(steps ~ date, activity, sum, na.rm = TRUE)

# shows the head of the steps_taken dataset
head(steps_taken)
```

Now we can plot the histogram from `steps_taken` to show the **total** number of steps taken per day using the appropriate bin interval.

```{r steps_taken_histogram}
# plot of total steps taken
g <- ggplot(steps_taken, aes(x=steps)) + 
        geom_histogram(fill=heat.colors(1), col="black", binwidth=1000) +
        labs(x = "Number of Steps per Day") + 
        labs(y = "Number of Times per Day") +
        labs(title = "Histogram of Total Number of Steps Taken per Day") +
        theme_bw()
print(g)
```

Now we can calculate the **mean** and **median** total number of steps taken per day.

```{r calculate_mean&median}
# calculate average mean and medium of steps taken
steps_taken %>% summarise(steps_mean= mean(steps), steps_median = median(steps))
```

1. The **mean** number of steps is: **10766.19**
2. The **median** number of steps is: **10765**

## What is the average daily activity pattern?

We will now plot a **time series** of the **average** number of steps taken across all days *(y-axis)*, of the 5-minute interval *(x-axis)*.

```{r interval_steps_taken_time_series}
interval_steps <- activity %>% group_by(interval) %>% summarise(avg_steps= mean(steps, na.rm=TRUE))

# time series plot of the 5-minute interval
g <- ggplot(interval_steps, aes(x=interval, y=avg_steps)) +
        geom_line(color="blue", size=1) +
        labs(title= "Average Daily Activity Pattern", x = "5-minute Interval", y="Number of Steps Taken") +
        theme_bw()
print(g)
```

We will now calculate which 5-minute interval, on average across all the days in **interval_steps** dataset, contains the **maximum** number of steps.
 
```{r maximum_steps}
interval_steps[which.max(interval_steps$avg_steps),]
```

We can see from the report that the **835^th^** 5-minute interval has an average maximum of **206** steps.

## Imputing missing values

We will now calculate and report the total number of rows with `NA`, from the **activity** dataset created earlier.

```{r calculate_NAs}
# using is.na() to calculate missing values.
sum(is.na(activity))
```

The total number of rows with **NAs** is: **2304**

Next we will create a new dataset called `fill_activity` by filling in all of the missing **NA** values in the `activity` dataset with the **mean** for 5-minute interval calculated in the `interval_steps` dataset.

```{r filling_dataset}
# make a copy of original dataset
fill_activity <- activity 
# make interval distinct
sapply(unique(activity$interval), 
       function(x) # function(x) is called to replace 'NA' values with mean 'values'
               fill_activity[is.na(fill_activity) & (fill_activity$interval == x), 1] <<- interval_steps$avg_steps[interval_steps$interval == x])
```

Even though the report shows no missing values in the dataset let us reaffirm.

```{r confirm_findings}
# using is.na() to calculate missing values.
sum(is.na(fill_activity))
```

The total number of rows with `NA` is: **0**

We will now plot an histogram of the total number of steps taken each day from the filled-in `fill_activity` dataset.

```{r filled-in-steps_taken-histogram}
fill_steps_taken <- aggregate(steps ~ date, fill_activity, sum, na.rm = TRUE)

# use the ggplot to make the hisgram of the total numbers of steps each day  
g <- ggplot(fill_steps_taken, aes(x=steps)) + 
        geom_histogram(fill=heat.colors(1), col="black", binwidth=1000) +
        labs(x = "Number of Steps per Day") + 
        labs(y = "Number of Times per Day") +
        labs(title = "Histogram of Total Number of Steps Taken per Day with Filled-in Data") +
        theme_bw()
print(g)
```

We will now calculate and report the **mean** and **median** total number of steps taken per day in the dataset `fill_steps_taken`.

```{r filled_compute_steps_taken}
fill_steps_taken %>% summarise(steps_mean= mean(steps), steps_median = median(steps))
```

1. The **mean** number of steps is: **10766.19**
2. The **median** number of steps is: **10766.19** 

We can see that the values of both the calculated *mean** and **median** of the filled-in dataset are equal to **10766.19**.

We can also see that the **mean = 10766.19** is the same for calculated steps in the dataset with missing values and dataset with filled-in values, while the **median** is showing a slight difference of **1.19** in favor of the dataset with filled-in values. 


```{r filled-interval_steps}
# dataset of filled-in values for the average 5-minute interval of steps taken
fill_interval_steps <- fill_activity %>% group_by(interval) %>% summarise(fill_avg_steps= mean(steps, na.rm=TRUE))

#calculated 5-minute interval on average of maximum number of steps from filled-in dataset
fill_interval_steps[which.max(fill_interval_steps$fill_avg_steps),]
```

Even though this calculation is not requested but is done to help confirm that if the **835^th^** 5-minute interval has the same average maximum of **206** steps for the filled-in dataset as for the missing dataset, then the impact of imputting missing data in the missing dataset remains the same for the maximum average numbers of steps. 


## Are there differences in activity patterns between weekdays and weekends?

We will next create a new factor variable in the `fill_activity` dataset with two levels - "weekday" and "weekend" indicating whether a given date is a `weekday` or `weekend` day.

```{r two_levels_variable}
fill_activity <- fill_activity %>%  
        mutate(days = as.factor(ifelse(weekdays(as.Date(date)) %in% c("Saturday","Sunday"), "Weekend", "Weekday"))) %>%
        group_by(days)
```

Next we will make a panel plot containing a time series plot of the 5-minute interval on the (x-axis) and the average number of steps taken, averaged across all `weekday` days or `weekend` days on the (y-axis), from the a dataset called `week_days`.

```{r days_time_series_plot }
week_days <- fill_activity %>%
        group_by(interval, days) %>%
        summarise(avg_steps = mean(steps))

g <- ggplot(week_days, aes(x=interval, y=avg_steps)) +
        geom_line(color="blue") +
        facet_wrap(~ days, nrow=2, ncol=1) +
        labs(x="interval", y="Number of Steps") +
        theme_bw()
print(g)
```

We observe from the graph above that more activities are done across all `weekends` than on `weekdays`. The maximum number of activities on average however, are between **500** and **1000** intervals and are done on `weekdays`. This could be that `steps` recorded on `weekdays` between these intervals are the result of a consistent activity pattern.

The rest of the intervals show on average, a higher but constant distribution of `activity` done on `weekends` than on `weekdays` and once again can be based on the result of a consistent activity pattern, that gradually decreases as the activity monitoring period comes to an end.

## Conclusion

This report concludes by asking; what would have been the results of the activity monitoring patterns in all required analysis, if the original dataset did not consist of **2304** missing values?